---
title: "Serum vs ABG electrolyte Comparison"
author: "Anupam kumar Singh , Imran Khan"
date: "26 October 2017"
output:
  pdf_document: default
  html_document: default
---

```{r,include=FALSE}
library(tidyverse)
electrolytes=read.csv("electrolytes.csv")
```

We Aim to analyse Following aspects in this article

1. Distribution of serum and point of care(POC) electrolytes variables
2. Correlation of serum and POC elecyrolytes
3. t. test of serum and POC electrolytes
4. Any misclassification of electrolyte categories by POC vis a vis serum test
5. Deming Regression , Normal regression , Regression equation of serum from POC test
6. Bland -Altman test,graphs, n stats of the two methods

All samples are independent in our article with no repeated measures for any patient

Let us start by creating categories of hypo/hyper/normo for sodium and potassium for both groups

```{r,echo=FALSE}
electrolytes= electrolytes %>% mutate(sodium_serum = case_when(
  serum_na <135 ~ "Hyponatremia",
  serum_na >150 ~ "Hypernatremia",
  TRUE ~ "Normal"
),sodium_poc = case_when(
  poc_na <135 ~ "Hyponatremia",
  poc_na >150 ~ "Hypernatremia",
  TRUE ~ "Normal"
) , potassium_serum = case_when(
  serum_k <3.5 ~ "Hypokalemia",
  serum_k >5.0 ~ "Hyperkalemia",
  TRUE ~ "Normal"
), potassium_poc = case_when(
  poc_k <3.5 ~ "Hypokalemia",
  poc_k >5.0 ~ "Hyperkalemia",
  TRUE ~ "Normal"
))


```





```{r,include=FALSE}
electrolytes= electrolytes %>%  mutate(potassium_serum=factor(potassium_serum,levels=c("Hypokalemia","Normal","Hyperkalemia")),potassium_poc=factor(potassium_poc,levels=c("Hypokalemia","Normal","Hyperkalemia")),sodium_poc=factor(sodium_poc,levels=c("Hyponatremia","Normal","Hypernatremia")),sodium_serum=factor(sodium_serum,levels=c("Hyponatremia","Normal","Hypernatremia")))
```

Let us see summary of our data

```{r,echo=FALSE}
electrolytes %>% select(poc_na,poc_k,serum_na,serum_k,potassium_serum,sodium_serum,potassium_poc,sodium_poc) %>% summary()
```

We can see that there are no hpernatremia patients in our population and misclassification is frequent in potassium group.

Let us visualise 

```{r,echo=FALSE}
electrolytes %>% ggplot(aes(x=poc_na))+geom_histogram()

electrolytes %>% ggplot(aes(x=serum_na))+geom_histogram()
electrolytes %>% ggplot(aes(x=poc_k))+geom_histogram()

electrolytes %>% ggplot(aes(x=serum_k))+geom_histogram()





```



Let us visualise comparative frequency polygons to have relative idea of distribution of sodium

```{r,include==FALSE}
electrolytes %>% select(poc_na,serum_na) %>% gather(key="samples",value = "value") %>% ggplot(aes(x=value,color=samples))+geom_freqpoly()
```

Let us visualise boxplot of sodium
```{r,echo=FALSE}
electrolytes %>% select(poc_na,serum_na) %>% gather(key="samples",value = "value") %>% ggplot(aes(y=value,x=samples))+geom_boxplot()
```

Let us visualise boxplot of potassium
```{r,echo=FALSE}
electrolytes %>% select(poc_k,serum_k) %>% gather(key="samples",value = "value") %>% ggplot(aes(y=value,x=samples))+geom_boxplot()
```

we can see serum potassium is slightly higher than poc potassium.

Now let us visualise scatterplot with regression line for sodium and potassium

```{r,echo=FALSE}
electrolytes %>% ggplot(aes(x=serum_na,y=poc_na))+geom_point()+geom_smooth(method="lm",formula = y~x)
```

```{r,echo=FALSE}
electrolytes %>% ggplot(aes(x=serum_k,y=poc_k))+geom_point()+geom_smooth(method="lm",formula = y~x)
```


Let us look at classification in potassium

```{r,echo=FALSE}
electrolytes %>% ggplot(aes(potassium_serum, ..count..)) + geom_bar(aes(fill = potassium_poc), position = "dodge")
```

Let us look at classification of sodium
```{r}
electrolytes %>% ggplot(aes(sodium_serum, ..count..)) + geom_bar(aes(fill = sodium_poc), position = "dodge")
```

Let us look at heat maps of sodium

```{r,echo=FALSE}
electrolytes %>% 
  count(sodium_serum,sodium_poc) %>%  
  ggplot(mapping = aes(x = sodium_serum, y = sodium_poc)) +
    geom_tile(mapping = aes(fill = n))
```

Let us look at heat maps of potassium

```{r}
electrolytes %>% 
  count(potassium_serum,potassium_poc) %>%  
  ggplot(mapping = aes(x = potassium_serum, y = potassium_poc)) +
    geom_tile(mapping = aes(fill = n))
```

Let us look at individual counts
```{r}
electrolytes %>% 
  count(sodium_serum,sodium_poc)
```

classification count for potassium

```{r}
electrolytes %>% 
  count(potassium_serum,potassium_poc) 
```



Now we have visualised data so it is time for some formal statistical tests, 
First test of skewness

```{r,include=FALSE}
library(e1071)
```


```{r}
electrolytes %>% select(serum_na,serum_k,poc_na,poc_k) %>% map(~skewness(.))
```

We see sodium data is relatively skewed 
```{r}
cor.test(electrolytes$poc_na,electrolytes$serum_na,method = "spearman")
```

Next Correlation test for potassium, We have used spearman's rho rank correlation due to non-normal distribution

```{r}
cor.test(electrolytes$poc_k,electrolytes$serum_k)
```

We can see correlation in potassium is a bit lower than correlation in potassium

Now time for t test for potassium we will use normal paired t test while for sodium wilcox.test due to its non-normal distribution

```{r}
wilcox.test(electrolytes$poc_na,electrolytes$serum_na,paired = TRUE)
```

we see non-significant difference in sodium

```{r}
t.test(electrolytes$poc_k,electrolytes$serum_k,paired = TRUE)
```
we see that serum potassium is signifcantly higher than poc potassium, notably it was obvious from boxplot as well ..now we have a statistical test to say the same

Now let us look at regression equation for serum sodium and potassium to see if they can be predicted from poc test

First for potassium

```{r}
f = lm(serum_k~poc_k,data=electrolytes)
summary(f)
```

Next for sodium

```{r}
fna = lm(serum_na~poc_na,data=electrolytes)
summary(fna)
```

we see adjusted R square is 0.92 for sodium and 0.81 for potassium implying POC sodium has higher predictive value for serum compared to potassium

We would also like to use deming regression for this analysis as it accounts for error in both variables and is required in clinical chemistry.
you can find details [here](https://en.wikipedia.org/wiki/Deming_regression) and [here](https://www.r-bloggers.com/deming-and-passing-bablok-regression-in-r/) 

```{r,include=FALSE}
library(mcr)
```

First Deming regression for sodium

```{r}

dem.sodium <- mcreg(electrolytes$serum_na,electrolytes$poc_na,method.reg = "Deming")
dem.sodium@para
```

Next Deming regression for Potassium

```{r}
dem.potassium <- mcreg(electrolytes$serum_k,electrolytes$poc_k,method.reg = "Deming")
dem.potassium@para
```

Deming regressions are of historical importance, but important because sometimes journals can ask for them

Now let us look at classification accuracy by kappa measurement. You can read more about it [here](https://en.wikipedia.org/wiki/Cohen%27s_kappa)

```{r,include=FALSE}
library(psych)

```

Cohen's kappa for Sodium

```{r}
elec_sod =electrolytes %>% select(sodium_poc,sodium_serum)
cohen.kappa(elec_sod)
```

Cohen's kappa for Potassium

```{r}
elec_pot =electrolytes %>% select(potassium_poc,potassium_serum)
cohen.kappa(elec_pot)
```

As expected cohen's kappa is slightly lower for potassium than sodium in conconcordance with our observations till now.


Now we will calculate Bland Altman Plot and statistics for sodium and potassium . You can read more about it in this [article](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4470095/) and this [article](https://www.medcalc.org/manual/blandaltman.php)


```{r,include=FALSE}
library(MethComp)
```

Bland Altman Plot of Sodium

```{r}
 with(electrolytes,BlandAltman(serum_na,poc_na))
```

Thus For sodium Serum sodium is higher than POC test by 0.36 and 95% C.I is likely to be between - 5.93 to 6.66 , It shoud be kept in mind that limit of allowable bias for sodium is 4 meq/L

Now Bland Altman Plot of Potassium

```{r}
 with(electrolytes,BlandAltman(serum_k,poc_k))
```

Thus For potassium Serum potassium is higher than POC test by 0.24 and 95% C.I is likely to be between - 0.57 to 1.0 , It shoud be kept in mind that limit of allowable bias for sodium is 0.5 meq/L
 
 